---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# elomakeR

Elo ratings for arbitrary head-to-head sports, built to interface with
the tidyverse. Column mapping uses tidy evaluation, so `calculate_elo()`
works directly with whatever schema your games data already has -- no
renaming required.

## Installation

Not (yet) on CRAN. From GitHub:

```r
# install.packages("devtools")
devtools::install_github("mattwaite/elomakeR")
```

## Example: 2025-26 NCAA D-I men's basketball

The bundled `basketball_games` dataset (pulled with `hoopR`) has
separate home/away columns and a real `neutral_site` flag for
tournament games played at neither team's venue:

```{r example}
library(elomakeR)
data(basketball_games)
head(basketball_games[, c("date", "home_location", "away_location", "home_score", "away_score", "neutral_site")])
```

`calculate_elo()` maps these columns directly, including the neutral
flag:

```{r calculate}
elo <- calculate_elo(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site
)
elo
```

The result is a long tibble -- one row per team per game -- which plugs
straight into dplyr/ggplot2 workflows. Current standings:

```{r standings}
latest_ratings(elo)
```

Rating trajectories for a few teams:

```{r plot, fig.width=7, fig.height=4}
library(ggplot2)
autoplot(elo, teams = c("Duke", "Michigan", "Houston"))
```

## Tuning the learning rate and home advantage

Rather than guessing `k` and `home_advantage`, search for the values
that best predict this season's (or a prior season's) results, scored
by log-loss:

```{r tune}
suggest_learning_rate(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site,
  k_grid = seq(10, 40, by = 5)
)
```

```{r tune-ha}
suggest_home_advantage(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site,
  home_advantage_grid = seq(0, 150, by = 25)
)
```

`tune_elo()` searches both parameters jointly when they might interact.
If the best value in a grid search lands on the edge of the grid you
passed in, widen the grid and try again -- that's a sign the true
optimum is further out.

## Margin of victory

Set `mov_adjustment = "log"` to scale the rating change up for lopsided
wins and down for narrow ones (ties are left untouched):

```{r mov}
calculate_elo(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site, mov_adjustment = "log"
) |>
  latest_ratings()
```

## Multi-season carryover

For data spanning multiple seasons, map a `season` column and choose
how ratings carry over at each season boundary: `"continuous"`
(default), `"reset"` (back to `initial_rating`), or `"regress"` (move a
fraction of the way back to `initial_rating`, via `regress_pct`).

## A messier schema: 2025 NCAA D-I women's volleyball

`calculate_elo()` doesn't care how odd the input schema is. The bundled
`volleyball_sets` dataset uses arbitrary `team1`/`team2` labels rather
than home/away columns, points to the home team with a separate
`home_team` column (`team2`, here), and records the outcome as sets won
(`team1_sets_won`/`team2_sets_won`, best of 5) rather than a single
score:

```{r volleyball}
data(volleyball_sets)
calculate_elo(
  volleyball_sets,
  date = date, team1 = team1, team2 = team2,
  score1 = team1_sets_won, score2 = team2_sets_won,
  home_team = team2
) |>
  latest_ratings()
```

## Data prep `calculate_elo()` won't do for you

`calculate_elo()` will happily rename columns for you, but it assumes
your data is already **one row per game**, with a single value each
for `team1`, `team2`, `score1`, and `score2` on that row. Column
mapping flexibility doesn't help if the rows themselves aren't shaped
that way yet -- and a lot of real schedule data isn't, because it's
built to answer "what did this team's season look like" rather than
"what happened in this game."

A concrete case: an NCAA baseball schedule scrape (one row per team per
game, not one row per game) that looks roughly like this:

```
team_name  opponent_name  team_score  opponent_score  game_type  result  contest_id
UCLA       UC San Diego   8           4               Home       W       6534937
UC Irvine  Sacramento St. 5           0               Home       W       6519201
```

Three mismatches with what `calculate_elo()` expects, and none of them
are things it should silently guess at -- getting them wrong doesn't
raise an error, it just produces wrong ratings:

- **Every game is listed twice** -- once from each team's perspective,
  same `contest_id`, scores mirrored. Feed both rows in and that game's
  rating update gets applied twice.
- **Home/away is per row, not per game.** A `game_type` column of
  `"Home"`/`"Away"`/`"Neutral"` describes *that row's* team, not a
  fixed home team for the game -- there's no single column you can
  point `home_team` at until you derive one.
- **Some rows have no result yet** (canceled, postponed, rained out)
  and no score to give `score1`/`score2`.

Clean all three with ordinary dplyr before calling `calculate_elo()`:

```r
library(dplyr)

games <- raw_schedule |>
  # drop games with no final score
  filter(result %in% c("W", "L", "T")) |>
  # collapse the two mirrored rows into one row per real game
  distinct(contest_id, .keep_all = TRUE) |>
  # turn the per-row home/away/neutral indicator into one home_team column
  mutate(
    home_team = case_when(
      game_type == "Home" ~ team_name,
      game_type == "Away" ~ opponent_name,
      TRUE ~ NA_character_
    ),
    neutral = game_type == "Neutral"
  )

calculate_elo(
  games,
  date = game_date, team1 = team_name, team2 = opponent_name,
  score1 = team_score, score2 = opponent_score,
  home_team = home_team, neutral = neutral, game_id = contest_id
)
```

A cheap sanity check that you deduplicated correctly: the output of
`calculate_elo()` always has exactly twice as many rows as its input
(one row per team per game), so `nrow(games) * 2` should equal
`nrow(elo)`. If a team's rating swings by an implausible amount after
one game, mirrored duplicate rows are the first thing to check for.
