---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# elomakeR

Elo ratings for arbitrary head-to-head sports, built to interface with
the tidyverse. Column mapping uses tidy evaluation, so `calculate_elo()`
works directly with whatever schema your games data already has -- no
renaming required.

## Installation

Not (yet) on CRAN or GitHub. From a local clone:

```r
# install.packages("devtools")
devtools::install(".")
```

## Example: 2025-26 NCAA D-I men's basketball

The bundled `basketball_games` dataset (pulled with `hoopR`) has
separate home/away columns and a real `neutral_site` flag for
tournament games played at neither team's venue:

```{r example}
library(elomakeR)
data(basketball_games)
head(basketball_games[, c("date", "home_location", "away_location", "home_score", "away_score", "neutral_site")])
```

`calculate_elo()` maps these columns directly, including the neutral
flag:

```{r calculate}
elo <- calculate_elo(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site
)
elo
```

The result is a long tibble -- one row per team per game -- which plugs
straight into dplyr/ggplot2 workflows. Current standings:

```{r standings}
latest_ratings(elo)
```

Rating trajectories for a few teams:

```{r plot, fig.width=7, fig.height=4}
library(ggplot2)
autoplot(elo, teams = c("Duke", "Michigan", "Houston"))
```

## Tuning the learning rate and home advantage

Rather than guessing `k` and `home_advantage`, search for the values
that best predict this season's (or a prior season's) results, scored
by log-loss:

```{r tune}
suggest_learning_rate(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site,
  k_grid = seq(10, 40, by = 5)
)
```

```{r tune-ha}
suggest_home_advantage(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site,
  home_advantage_grid = seq(0, 150, by = 25)
)
```

`tune_elo()` searches both parameters jointly when they might interact.
If the best value in a grid search lands on the edge of the grid you
passed in, widen the grid and try again -- that's a sign the true
optimum is further out.

## Margin of victory

Set `mov_adjustment = "log"` to scale the rating change up for lopsided
wins and down for narrow ones (ties are left untouched):

```{r mov}
calculate_elo(
  basketball_games,
  date = date, team1 = home_location, team2 = away_location,
  score1 = home_score, score2 = away_score,
  home_team = home_location, neutral = neutral_site, mov_adjustment = "log"
) |>
  latest_ratings()
```

## Multi-season carryover

For data spanning multiple seasons, map a `season` column and choose
how ratings carry over at each season boundary: `"continuous"`
(default), `"reset"` (back to `initial_rating`), or `"regress"` (move a
fraction of the way back to `initial_rating`, via `regress_pct`).

## A messier schema: 2025 NCAA D-I women's volleyball

`calculate_elo()` doesn't care how odd the input schema is. The bundled
`volleyball_sets` dataset uses arbitrary `team1`/`team2` labels rather
than home/away columns, points to the home team with a separate
`home_team` column (`team2`, here), and records the outcome as sets won
(`team1_sets_won`/`team2_sets_won`, best of 5) rather than a single
score:

```{r volleyball}
data(volleyball_sets)
calculate_elo(
  volleyball_sets,
  date = date, team1 = team1, team2 = team2,
  score1 = team1_sets_won, score2 = team2_sets_won,
  home_team = team2
) |>
  latest_ratings()
```
